---
title: "Introduction to R Programming"
subtitle: "The Apply Family"
author: "Pedro Fonseca"
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Built in datasets

R provides many built in datasets. For a complete list see:

```{r, eval = FALSE}
library(help = "datasets")
```

In this lecture we will use some of these datasets, namely:

  - USArrests
  - airquality

For information about a specific dataset see, for example:

```{r, eval = FALSE}
?USArrests
```

## The `USArrests` dataset

- This data set contains the arrests per 100.000 residents for assault, murder, and rape in each of the 50 USA states in 1973.

- There is also one column with the percentage of population living in urban areas.

## The `USArrests` dataset

```{r}
head(USArrests)
```

## The `USArrests` dataset

Let's do some calculations with the arrests of four states:

```{r}
USArrests_short <- USArrests[1:4, -3]
                            
USArrests_short
```

## Row and column sums

```{r}
colSums(USArrests_short) 
rowSums(USArrests_short) 
```

## Row and column means

```{r}
colMeans(USArrests_short) 
rowMeans(USArrests_short) 
```

## The `apply()` function

To collapse data frames across rows or columns using functions other than the sum and the mean we can use `apply()`:

`apply(X, MARGIN, FUN, ...)`

  - X is a data frame
  - `MARGIN = 1` for rows, `MARGIN = 2` for columns
  - FUN is a function
  - ... are optional arguments to pass to FUN

## The `apply()` function

Apply functions over the columns of `USArrests_short`:

```{r}
apply(USArrests_short, 2, mean)
apply(USArrests_short, 2, median)
apply(USArrests_short, 2, sd)
```

## The `apply()` function

Apply functions the rows of `USArrests_short`:

```{r}
apply(USArrests_short, 1, max)
apply(USArrests_short, 1, min)
apply(USArrests_short, 1, var)

```

## `apply()` vs `for` loop

Loop over the columns of `USArrests_short`:

```{r}
res <- vector()

for(i in 1:ncol(USArrests_short)){
  
  res[i] <- mean(USArrests_short[[i]], na.rm = TRUE)
  names(res)[i] <- names(USArrests_short)[i]
}

res
```

## `apply()` vs `for` loop

Loop over the rows of `USArrests_short`:

```{r}
res <- vector()

for(j in 1:nrow(USArrests_short)){
  
  res[j] <- max(USArrests_short[j, ],na.rm = TRUE)
  names(res)[j] <- rownames(USArrests_short)[j]
}

res
```

## `apply()` with `...`

Now let's see an example that requires using dot-dot-dot (`...`). Suppose you're working with the `airquality` the data frame:

```{r}
head(airquality)
```

## `apply()` with `...`

Try to use `apply()` to compute the means of the first four columns:

```{r}
apply(airquality[, 1:4], 2, mean)
```

We get `NA` for the first two columns. Why?

## `apply()` with `...`

Problem: Some columns have `NA`s.

```{r}
sum(is.na(airquality$Ozone))
sum(is.na(airquality$Solar.R))
```

## `apply()` with `...`

Solution: use `...` to pass the `na.rm` argument to `mean()`:

```{r}
apply(airquality[, 1:4], 2, mean, na.rm = TRUE)
```

## `apply()` user-defined functions

Checking for `NAs` one column at a time is not be feasible when we have many columns. How can we check how many `NA`s there there in each column?

```{r}
count_nas <- function(x){sum(is.na(x))}

apply(airquality, 2, count_nas)
```

## The `lapply()` function

Syntax: `lapply(X, FUN, ...)`

`lapply()` is similar to `apply()` but:

  - X is a vector (atomic or list)
  - There is no MARGINS argument
  - Always returns a list
  
`lapply()` returns a list of the same length as X, each element of which is the result of applying FUN to the corresponding element of X.

## The `lapply()` function

```{r}
lapply(c(1:3), log, base = 10)
```

## The `lapply()` function

```{r}
A <- matrix(1:10, ncol = 5)
B <- matrix(c(1, 5, 7, -1), ncol = 4)
C <- matrix(letters[1:4], ncol = 2)

my_list <- list(A, B, C)

```

## The `lapply()` function

```{r}
my_list
```

## The `lapply()` function

Every element of `my_list`, except the last, contain a numerical matrix. Sum the elements of each of those matrices:

```{r}
lapply(my_list[-3], sum)
```

## The `lapply()` function

Extract the element in position (1, 2) from each matrix:

```{r}
lapply(my_list,"[", 1, 2)
```

## The `lapply()` function

Extract the first row from each matrix:

```{r}
lapply(my_list,"[", 1 , )
```

## The `lapply()` function

Extract the 2nd column from each matrix:

```{r}
lapply(my_list,"[", , 2)
```

## `lapply()` vs `for` loop

Extract the 2nd column from each matrix:

```{r}

res <- vector(mode = "list")

for(i in seq_along(my_list)){
  res[[i]] <- my_list[[i]][, 2]
}
res
```

## The `sapply()` function

The sapply() function:

  - Works like lapply(), but simplifies the output to the most elementary data structure that is possible.
  - Returns vectors or matrices.

## The `sapply()` function

```{r}
sapply(my_list[-3], sum)
sapply(my_list,"[", 1, 2)
```

## The `sapply()` function

```{r}
sapply(airquality, function(x) sum(is.na(x)))
```

## The `sapply()` function

```{r}
set.seed(123)

our_list <- list(
  w = 1:6,
  x = sample(1:5, 4, replace = TRUE),
  y = matrix(sample(1:100, 9), nrow = 3),
  z = sample(1:10, 3, replace = TRUE) 
)

str(our_list)
```

## The `sapply()` function

```{r}
sapply(our_list, max)
sapply(our_list, min)
sapply(our_list, class)
```

## The `sapply()` function

How many numbers are there inside each element of `our_list`?

```{r}
sapply(our_list, length)
```

## The `sapply()` function

```{r}
set.seed(123)

our_list_2 <- list(
  w = 1:6,
  x = sample(1:5, 4, replace = TRUE),
  y = airquality
)
```

```{r}
sapply(our_list_2, class)
dim(airquality)
```

How many numbers are there inside each element of `our_list_2`? 

## The `sapply()` function

The length of a data frame is the number of columns, and hence `sapply(our_list_2, length)` won't do the trick.

```{r}
our_fun <- function(x){
  if(class(x) == "data.frame"){
    nrow(x) * ncol(x)
  }else{
      length(x)
  }
}

sapply(our_list_2, our_fun)
```

## `sapply()` vs `for` loop

The same but with a `for` loop:

```{r}

res <- vector()

for(i in seq_along(our_list_2)){
  
  res[i] <- our_fun(our_list_2[[i]])
  names(res)[i] <- names(our_list_2)[i] 
}

res

```

## The `mapply()` function

`mapply()` is a generalization of `sapply()`.  It applies a multivariate function over multiple vectors of arguments.

## The `mapply()` function

Suppose we want 3 samples of different sizes from a $\operatorname{Normal}(0, 1)$ distribution:

```{r}
set.seed(123)

rnorm(n = 1)
rnorm(n = 2)
rnorm(n = 3)
```

## The `mapply()` function

The same result can be obtained more compactly with `mapply()`:

```{r}
set.seed(123)

sample_size <- 1:3
mapply(FUN = rnorm, n = sample_size)
```

## The `mapply()` function

Since we only iterated over one vector, we could have used `sapply()`:

```{r}
set.seed(123)

sample_size <- 1:3
sapply(FUN = rnorm, X = sample_size)
```

## The `mapply()` function

But what if we want to sample from normal distributions with different means, while still having samples of different sizes? 

```{r}
set.seed(123)

rnorm(n = 1, mean = 5)
rnorm(n = 2, mean = 10)
rnorm(n = 3, mean = -3)
```

## The `mapply()` function

In this case we need to iterate over two vectors, one for sample sizes and one for means:

```{r}
set.seed(123)

sample_size <- 1:3
mu <- c(5, 10, -3)
mapply(rnorm, n = sample_size, mean = mu)

```

## The `mapply()` function

Now suppose we also want each sample to have a different standard deviation:

```{r}
set.seed(123)

rnorm(n = 1, mean = 5, sd = 1)
rnorm(n = 2, mean = 10, sd = 3)
rnorm(n = 3, mean = -3, sd = 5)
```

## The `mapply()` function

```{r}
set.seed(123)

sample_size <- 1:3
mu <- c(5, 10, -3)
sigma <- c(1, 3, 5)

mapply(rnorm, mean = mu, sd = sigma, n = sample_size)
```

## The `mapply()` function

Now suppose we wanted our results with two decimal places only:

```{r}
set.seed(123)

results <- mapply(rnorm, mean = mu, sd = sigma,
                  n = sample_size)
```

```{r}
sapply(results, FUN = round, 2)
```

## `mapply()` vs `for` loop

```{r}
set.seed(123)

sample_size <- 1:3
mu <- c(5, 10,-3)
sigma <- c(1, 3, 5)

res <- vector(mode = "list")

for (i in 1:3) {
  res[[i]] <- round(rnorm(mean = mu[i],
                          sd = sigma[i],
                          n = sample_size[i]),
                    2)
}
```

## `mapply()` vs `for` loop

```{r}
res
```
