---
title: "Introduction to R Programming"
subtitle: "Lists"
date: "`r format(Sys.time(), '%d %B %Y')`"
author: Pedro Fonseca
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminars

The `str()` function displays the internal strucure of an object:

```{r}
vec <- c(1, 5, 0.1)
str(vec)

vec_f <- factor(c("good", "bad"))
str(vec_f)

mat_char <- matrix(c("a", "u", "mk", "q!"), ncol = 2)
str(mat_char)
```

## Preliminars

![The hierarchy of R’s vector types](vec){width=250px}

## Why do we need lists?

- Atomic vectors and matrices are convenient data storage structures but they have a limitation: all their elements have to be of the same data type.

- If we provide elements of more than one data type to an atomic vector or matrix, R will coerce to the most general data type.

## Why do we need lists?

```{r}
vec1 <- c(TRUE, FALSE)
typeof(vec1)
```

```{r}
vec2 <- c(TRUE, FALSE, 3)
typeof(vec2)
```

Logical values are coerced to numeric if combined with numeric values.

## Why do we need lists?

```{r}
vec3 <- c(TRUE, FALSE, 3, "4")
typeof(vec3)
```

Logical and numeric values are coerced to character if combined with character values.

## What are lists?

- Lists are a step up in complexity from atomic vectors.

- Atomic vectors are homogeneous, while lists can be heterogeneous. 

- Lists can store objects of any type (including lists).

- Lists are recursive vectors.


## Creating a list

```{r, echo = TRUE}
my_list <- list(
  matrix(c(1, 5, -1, 4), ncol = 2),
  c("S", "M", "S", "W"),
  "I-am-a-string",
  c(TRUE, FALSE, FALSE)
)
```

```{r}
str(my_list)
```

## Whats inside the list?

```{r}
my_list
```

## Lists inside lists

Lists can even contain other lists!

```{r}
list_with_lists <- list(
  c(4:1),
  list(
    matrix(c("d", "a", "r", "r"), ncol = 2),
    factor("yes", "no", "yes")
    ),
  list(
    c(TRUE, FALSE, TRUE),
    1:4,
    c(1, 5, 7, 9, 0)
    )
  )

```

## Lists inside lists

```{r}
str(list_with_lists)

```

## Lists inside lists inside lists...

```{r}
a_list_of_lists <- list(list(list(list(c(1, 5),
                                       c("a", "b")))))
str(a_list_of_lists)
```

This is why lists are said to be recursive.

## Naming list elements

```{r}
names(my_list) <- c("Numbers", "Letters",
                    "a_lonely_string", "T_F")
str(my_list)
```

## Naming list elements

```{r}
my_list
```

## Naming list elements

We can also name the elements of the list when we create it:

```{r, echo = TRUE}
my_list <- list(
  some_numbers = matrix(c(1, 5, -1, 4), ncol = 2),
  some_letters = c("S", "M", "S", "W"),
  a_lonely_string = "I-am-a-string",
  T_or_F = c(TRUE, FALSE, FALSE)
)

str(my_list)
```

## Naming list elements

`list()` does not keep the names of input objects:

```{r}
num_vec <- 1:3
char_mat <- matrix(c("a", "b", "c", "m"), ncol = 2)
a_lonely_string <- "Hello!"
a_factor <- factor(c("Yes", "No", "Yes"))

my_list_2 <- list(
  num_vec,
  char_mat,
  a_lonely_string,
  a_factor,
  my_list 
  )
```

## Naming list elements

```{r}
str(my_list_2)
```

## Naming list elements

```{r}
num_vec <- 1:3
char_mat <- matrix(c("a", "b", "c", "m"), ncol = 2)
a_lonely_string <- "Hello!"
a_factor <- factor(c("Yes", "No", "Yes"))

my_list_2_named <- list(
  numbers = num_vec,
  c_mat = char_mat,
  lonely = a_lonely_string,
  fac = a_factor,
  old_list = my_list 
  )
```

## Naming list elements

```{r}
str(my_list_2_named)
```

## Coercion 

We can turn objects into a lists with `as.list()`: 

```{r}
char_vec <- c("yes", "no")
as.list(char_vec)
```

```{r}
str(as.list(char_vec))
```

## Coercion 

```{r}
num_matrix <- matrix(1:4, ncol = 2)
as.list(num_matrix)

```

## Subsetting

We can subset lists with:

- The `[` operator. Always returns a list. 
- The `[[` operator. Returns the object that is inside the subsetted element of the list. Can return objects of any type.
- The `$` operator. Simillar to `[[` but works only with named list elements. 

## Subsetting

```{r}
str(my_list)
```

```{r}
my_list[1] 
typeof(my_list[1])
```

## Subsetting

```{r}
is.matrix(my_list[1])
my_list[[1]]
typeof(my_list[[1]])
is.matrix(my_list[[1]])
```

## Subseting

> "If list x is a train carrying objects, 
>  then x[[5]] is the object in car 5; 
>  x[4:6] is a train of cars 4-6" 
> `r tufte::quote_footer('--- @RLangTip')`

## Subsetting 

If element names are available:

```{r}
my_list["some_letters"]
```

```{r}
my_list[["some_letters"]]
```

```{r}
my_list$some_letters
```

## Subsetting

Because it can return only a single value, you must use `[[` with either a single positive integer or a string:

```{r}
a_short_list <- list(a = c(1, 2), b = 3)
a_short_list[[1]]
a_short_list[["a"]]
```

## Subsetting

If you do supply a vector to `[[` it indexes recursively:

```{r}
rec_l <- list(a = list(b = list(c = list(d = 3))))
str(rec_l)

rec_l[["a"]][["b"]][["c"]][["d"]] 

rec_l[[c("a", "b", "c", "d")]] # Same as above!

```

## Subsetting

Some examples using `my_list`:

```{r}
my_list
```

## Subsetting

```{r}
my_list[c(2, 3)] 
```

```{r}
typeof(my_list[c(2, 3)]) 
length(my_list[c(2, 3)])
```

## Subsetting

Extracting an elements from a vector that is inside a list:

```{r}
my_list[[2]][1]
```

or:

```{r}
my_list[[c(2, 1)]]
```

Extracting elements of a vector that is inside a list:

```{r}
my_list[[2]][c(2, 3)]
```

## Subsetting

```{r}
my_list[[1]]
```

Extracting elements of a matrix that is inside a list:

```{r, eval = FALSE}
my_list[[1]][, 2] # Extracts the second column
my_list[[1]][1, ] # Extracts the first row
my_list[[1]][-1, ] # Matrix without the first row
my_list[[1]][1, 1] # Element in position (1,1)
```

## Add a new element to a list using `[[`

```{r}
length(my_list)

my_list[[5]] <- matrix(c(0.21, 0.45, 0.6, 3),
                       ncol = 2)

length(my_list)
```

Note that this new element is not named. Using `[` instead of `[[` would also work.

## Add a new element to a list using `$`

Adding elements to a list with `$` automatically names the new element.

```{r}
my_list$new_thing <- list(c(1, 5), "some-stuff", 
                        matrix(1:6, nrow = 2))
str(my_list)

```

## Extracting elements from a list inside a list

Element 6 of `my_list` is also list:

```{r}
str(my_list[6])
```

## Extracting elements from a list inside a list

Extract the element in the position (2, 3) of the matrix that is inside the list that is itself inside `my_list`:

```{r}
my_list[[6]][[3]][2, 3]
```

Since the list that inside `my_list_2` is named, this also works:

```{r}
my_list[["new_thing"]][[3]][2, 3]
```

## Delete a element of a list

```{r}
str(my_list)
```

## Delete an element of a list

`NULL` is often used to represent the absence of a vector (as opposed to `NA` which is used to represent the absence of a value in a vector). 

```{r}

my_list$new_thing <- NULL
# my_list["new_thing"] <- NULL does the same

str(my_list)
```

## Delete an element of a list

```{r}
my_list[[5]] <- NULL
str(my_list)
```

## Delete an element of a list

```{r}
my_list["a_lonely_string"] <- NULL
# my_list[2] <- NULL does the same

str(my_list)
```

## Merge lists

You can merge lists with `c()`:

```{r}
list1 <- list(1, 2)
list2 <- list(c("Sun","Mon"), c(1, 2))
list3 <- list("Hi!")

merged.list <- c(list1, list2, list3)
str(merged.list)
```

## Unlist

```{r}
l <- list(
  x = 1:2,
  y = c("ab", "3"),
  m = matrix(1:4, ncol = 2)
)

unlist(l)
```

The `unlist` function:

- Keeps names (if available)
- Collapses a list to a vector
- Coerces to the most general data type
