---
title: "Introduction to R Programming"
subtitle: "Getting Started"
author: Pedro Fonseca
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R and Rstudio

- R is a programming language and free software environment for statistical computing and graphics.	
- RStudio is an integrated development environment (IDE) for R.
- You can use R without using RStudio, but you can't use Rstudio without using R.

## This is how R looks like

\begin{figure}
\includegraphics[scale = .3]{figures/Screenshot_1}
\caption{R console on windows}
\end{figure}

## This is how R looks like

\begin{figure}
\includegraphics[scale = .35]{figures/r-mac.png}
\caption{R console on MacOS}
\end{figure}

## This is how R looks like

\begin{figure}
\includegraphics[scale = .3]{figures/r-linux}
\caption{R on Ununtu}
\end{figure}

## This is how Rstudio looks like

\begin{figure}[H]
\includegraphics[scale = 0.15]{figures/Screenshot_mac}%
\caption{Rstudio on MacOS}
\end{figure}

## Rstudio Cloud

If you don't want to install R and RStudio:		

1. Go to [RStudio Cloud](https://rstudio.cloud/)

2. Create an account and login

3. Click "New Project"

## Rstudio Cloud

\begin{figure}[H]
\includegraphics[scale = .3]{figures/Screenshot_6} 
\caption{Rstudio Cloud}
\end{figure}

For additional information see [https://rstudio.cloud/learn/guide](https://rstudio.cloud/learn/guide)

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.35]{figures/new-project-1.png}
\end{figure}

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.43]{figures/new-project-2.png}
\end{figure}

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.43]{figures/new-project-3.png}
\end{figure}

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.43]{figures/new-project-4.png}
\end{figure}

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.33]{figures/new-project-5.png}
\end{figure}

## Your first Rstudio project

\begin{figure}
\includegraphics[scale=0.19]{figures/environment.png}
\end{figure}

## R as a calculator

We can use R's console as a calculator:

```{r}
2+3 
3*5
14.5/6
3^2
```

## R as a calculator

```{r}
(3^2)+14/(6+5)
(3^2)+14/6+5
```

## R as a calculator

```{r}
25^0.5
sqrt(25)
```

## R as a calculator

```{r}
log(5)
log10(5)
```

## R as a calculator

```{r}
pi
cos(2*pi)
tan(0.6)
sin(0.6)/cos(0.6)
```

## Functions

- R has a large collection of built-in functions.

- We´ve already used `log`, `log10`, `sqrt`, `sin`, `cos` and `tan`

## Functions

This is how you call a function:

```{r, eval=FALSE}
function_name(arg1 = val1, arg2 = val2, ...)
```

- Some arguments are mandatory. 
- Some arguments are optional and have default values.
- Argument names are not mandatory. 
- If you don't provide the names of the arguments, you must input the arguments in the correct order.
- As long as the argument's names are provided, the order is irrelevant.
- Help pages can be useful.

## Getting Help

- If you don't know what a functions does just put \,"?"\, before the name of the function and send it to R's console. 

- In the help page a function you can find:
  
- Its arguments and respective admissible values
- The interpretation of its output
- Examples
- Related functions

```{r, eval=FALSE}
?mean
?library
?sqrt
```

## Some examples of functions

The exponential function is given by `exp()`.

```{r}
exp(x=3)
```

## Some examples of functions

Logarithms can be calculated with the `log` function.

```{r}
log(x = 243, base = 3)
log(x = 243)
```

The \textit{base} argument is optional. The default value is $e$.

## Some examples of functions

```{r}
log(243, exp(1))
log(exp(1), 243)
log(base = exp(1), x = 243)
```

Tip: try

```{r, eval=FALSE}
?log
```

## Some examples of functions

```{r}
log(x = 243, base = exp(1))
log10(5)
2^log2(6)
10^log10(5)+1
```

## R scripts

- We´ve been using R's console

- Code sent directly to the console is executed but you won't be able to modify it or reuse it later

- Writting our code in scripts is a better option

- A script is just a text file we can use to write code

## Your first R script

\begin{figure}
\includegraphics[scale=0.35]{figures/new-script.png}
\end{figure}

## Rstudio Panes

\begin{figure}
\includegraphics[scale=0.18]{figures/panes.png}
\end{figure}

## Editor

-  R opens scripts in the editor pane
-  This is where you should write your code
-  In the editor you can modify, rerun and save your code at any time

## Some Useful Shortcuts

- New script: Cmd/Ctrl + Shift + N 
- Save the script: Cmd/Ctrl + S
- Send code from the editor to the console:

  - Cmd/Ctrl + Enter (current line or current selection)
  - Cmd/Ctrl + Shift + S (entire script)

## More Shortcuts

- To see a list of Rstudio shortcuts try: Alt/Option + Shift + k
  
- Alternative: click [here](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)

## Assigning values to objects

To store values in R's memory you need to assign them to objects. You can use the equal sign, the assign function, or the assign operator:

- The assignment operator is typically recommended. 
- The equal sign should be reserved to provide arguments to functions. 

```{r}
object_name_1 <- 5
object_name_1

object_name_2 <- log(object_name_1) + exp(5)
object_name_2
```

Rstudio's keyboard shortcut for the assign operator: "Alt/Option" + "$-$"

## The assignment operator

\includegraphics[scale=.22]{figures/meme.jpg}

## The assignment operator

\begin{figure}
\includegraphics[scale=.19]{figures/objects.png}
\caption{Stored objects are visible in the upper-right pane, under the "Environment" tab}
\end{figure}

## Example

\begin{figure}
\includegraphics[scale=0.17]{figures/edit-script.png}
\end{figure}

## Example (cont.)

\begin{figure}
\includegraphics[scale=0.17]{figures/saved-script.png}
\end{figure}

## Naming Objects

Object names must start with a letter and can only contain letters, numbers, underscores and dots. You want your object names to be short, descriptive and consistent. Ideally, one should follow a convention:

- i_use_snake_case
- otherPeopleUseCamelCase
- some.people.use.periods
- And_aFew.People_RENOUNCEconvention

## Case Matters

```{r, error=TRUE}
pi
r_rocks <- 2 * pi^2
r_rocks
r_Rocks

```

## How to delete objects

To delete stored objects use the `rm` function:

```{r, error=TRUE}
rm(object_name_1)
object_name_1
```

## How to delete objects

- You can input as many objects as you want to `rm()`

- To remove all stored objects all once, use the following  command:
  
```{r, eval=FALSE}
rm(list = ls())
```

## How to print the assigned value

If you make an assignment, you don't get to see the assigned value. You're then tempted to double-check the result:
  
```{r}
y <- log(2)+1
y
```

This common action can be shortened by surrounding the assignment with parentheses, which causes assignments to print:

```{r}
(y <- log(2)+1)
```

## Overwritting stored  values

```{r}
x <- -5
x

x <- x + 1 
x
```

## Working directory

An active R session always has an associated working directory. R will use the working directory by default to:

- Search for files
- Save outputs (tables, plots, etc)

## Setting the working directory

\begin{figure}
\includegraphics[scale = .38]{figures/wd}
\caption{Setting the working directory}
\end{figure}

## Setting and setting the working directory

You can also get or set the working directory in R's console:

```{r, eval=FALSE}
getwd()
```

```{r, eval=FALSE}
setwd("/folder1/folder2/folder3/")
```

- The problem with commands like this is that such paths will only exist on your computer
- Solution: Rstudio projects

## Advantages of Rstudio projects

- Rstudio projects are self-contained.
- They put together all the files that are relevant for a particular project (article, book, research project) in the same folder.
- The project's working directory always points to that folder by default
- Rstudio projects can be moved around on your computer or onto other computers and will still “just work”. No directory changes are needed.
- If you need to create additional folders or start moving around parts of you project around dont use the `setwd` function. It is safer to reference the full path.

## Packages

- The more specialized functions and data sets are available on packages (also referred to as libraries).

- Installing R Packages:

```{r, eval=FALSE}
install.packages("ggplot2", dependencies = TRUE)
```

Loading R Packages:

```{r, eval=FALSE}
library("ggplot2")
```

Updating R Packages:

```{r, eval=FALSE}
update.packages()  # This is rarely necessary 
```

- Packages are developed by the R core team and also by the community of R users.
- You can develop your own packages and make them available to the community on [CRAN](https://cran.r-project.org)(The Comprehensive R Archive Network)

## Packages

- It is typically recommend to start your scripts with the packages that you need. 
- That way, if you share your code with others, they can easily see what packages they need to install. 
- Note, however, that you should never include `install.packages` or `setwd` in a script that you share. 
- It is very antisocial to change settings on someone else’s computer!

## Settings

You can change Rstudio's default settings and appearence:

- Mac: Tools $->$ Global Option
- Windows and Linux: Rstudio $->$ preferences

Shortcut:

- Mac: "Cmd" + "," 
- Windows and Linux: "Ctrl" + "," 

## Settings

\begin{figure}
   \includegraphics[scale = .3]{figures/settings.png}
   \caption{These are the general settings that we recommend}
\end{figure}

## Settings

\begin{figure}
   \includegraphics[scale = .35]{figures/Screenshot_4}
   \caption{Changing Rstudio's appearence}
\end{figure}

