---
title: "Introduction to R Programming"
subtitle: "Data Frames"
author: Pedro Fonseca
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
output: beamer_presentation
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Before we start

The `ifelse()` function performs vectorized if-then-else statements

```{r, eval = FALSE}
ifelse(test, x, y)
```

- `test` can be anything that returns a logical vector or matrix

- `ifelse()`returns `x` in the positions where the corresponding element of `test` is `TRUE` and `y` in the positions where the corresponding element of `test` is `FALSE`

## Before we start

```{r}
vec <- 1:5
```

```{r}
ifelse(vec > 2, 2, 1)
ifelse(vec > 2, "large", "small")
ifelse(vec > 2, vec - 1, vec + 1)
```

## Before we start

You can chain multiple `ifelse()` calls:

```{r}
heights <- c(1.66, 1.88, 1.76, 1.68, 1.7, 1.9)

shirts <- ifelse(heights >= 1.9, "XL",
                 ifelse(heights > 1.8, "L",
                        ifelse(heights > 1.75, "M",
                               "S")
                        )
                 )

shirts
```

## Datasets

In a typical dataset:

- Rows represent observations
- Columns represent variables

  
![A retangular dataset](figures/dataset){width=200px}

## What is a data frame?

Data frames are designed specifically for rectangular datasets

- Why not matrices? We may need different data types for different columns

- Why not lists? Not very practical nor visually appealing

## My first data frame

You can create a data frame by supplying name-vector pairs to `data.frame()`:

```{r}
df <- data.frame(
  age = c(20, 38, 50, 19),
  smoker = c(TRUE, FALSE, FALSE, FALSE),
  marital = factor(c("M", "S", "W", "S"))
)
```

## My first data frame

Let's print `df`:

```{r}
df
```

Notice that data frame rows are numbered by default

## Some utilities 

```{r}
summary(df)
```

Try this:

```{r, eval = FALSE}
View(df)
```

## What is a data frame?

```{r}
str(df)
```

- Similar to the structure of a list?
- Actually, data frames are build on top of lists
- A data frame is a list of atomic vectors with equal length

## What is a data frame?

Data frames have attributes for column `names`, `row.names`, and their own class: `data.frame`:

```{r}
typeof(df) 
class(df) 
```

## What is a data frame?

Due to their rectangular structure, data frames share the properties of both matrices and lists:

- A data frame has `rownames()` and `colnames()`. The `names()` of a data frame are the column names

- A data frame has `nrow()` rows and `ncol()` columns. `The length()` of a data frame is the number of columns

## My second data frame

```{r}
name <- c("Anne", "Pete", "Frank", "Julia", "Cath")
age <- c(28, 30, 21, 39, 35)
child <- c(FALSE, TRUE, TRUE, FALSE, TRUE)

df2 <- data.frame(name, age, child)
df2
```

Unlike `list()`, `data.frame()` keeps the names of the input vectors

## Assign `names()` to a data frame

Data frame columns always have `names`. If you don´t provide them,
default names will be assigned:

```{r}
df3 <- data.frame(
  c(1, 4, 6),
  c(6, 7, 9)
) 

df3
```

## Assign `names()` to a data frame

```{r}
names(df3) <- c("a-name", "some-other-name")

df3
```

## Row names

`data.frame()` allows you to label row indexes. You can provide a vector of unique values to the `row.names` argument:

```{r}
df4 <- data.frame(
  age = c(35, 27, 18),
  hair = c("blond", "brown", "black"),
  row.names = c("Bob", "Susan", "Sam") # row names!
) 

df4
```

## Row names

`rownames()` can be used to get, set or overwrite the row names of an existing data frame:

```{r}
rownames(df4)

rownames(df4) <- c("Bob M.", "Susan B.", "Sam L.") 

df4
```

## `stringsAsFactors`

```{r}

df5 <- data.frame(
  person = c("John", "Matt", "Mary"),
  age = c(20, 38, 50),
  smoker = c(TRUE, FALSE, FALSE),
  marital = factor(c("M", "S", "S"))
)

str(df5)
```

The `person` variable is a factor! Why?

## stringsAsFactors

The `stringsAsFactors` argument of `data.frame` is set to `TRUE` by default.

```{r}
df6 <- data.frame(
  person = c("John", "Matt", "Mary"),
  age = c(20, 38, 50),
  smoker = c(TRUE, FALSE, FALSE),
  marital = factor(c("M", "S", "S")),
  stringsAsFactors = FALSE
)
str(df6)
```

## Testing and coercing

To check if an object is a data frame use `is.data.frame()`:

```{r}
is.data.frame(df6)
```

You can coerce an object to a data frame with `as.data.frame()`:

```{r}
M <- matrix(1:9, ncol = 3)
as.data.frame(M)
```

## Subsetting

Data frames inherit subsetting syntax from both matrices and lists. You can subset a data frame:

- Like a 2D structure (behaves like a matrix)
- Like a 1D structure (behaves like a list)

## Subsetting

```{r}
name <- c("Anne", "Pete", "Frank", "Julia", "Cath") 
age <- c(28, 30, 21, 39, NA)
child <- c(FALSE, FALSE, TRUE, FALSE, TRUE)

people <- data.frame(name, age, child, 
                     stringsAsFactors = FALSE)
people
```

## Subsetting with matrix syntax

```{r}
people[3, ] # returns a data frame
people[c(1, 3), ] # data frame
people[3, 2] # vector
people[3, "age"] # vector

```

## Subsetting with matrix syntax

```{r}
people[, "age"] # vector
people[, 1] #  vector
people[c(3, 5), c("age", "child")] # data frame
```

## Subsetting with matrix syntax

See the pattern?

- When you subset a data frame like a 2D object with `[` you get a vector if only one column is extracted. You get a data frame otherwise.

- This is a frequent source of bugs when using `[` in a function, unless you always remember to use `drop = FALSE`.

## Subsetting with matrix syntax

```{r}
people[, "age", drop = FALSE] # data frame!
```

Setting `drop = FALSE` prevents `[` from coercing to a lower dimension.

## Subsetting with matrix syntax

You can also subset using logical conditions. Show Pete's row:

```{r}
people[people[, "name"] == "Pete", ]
```

## Subsetting with matrix syntax

Show Julia's age and wheather or not she has children:

```{r}
people[people[, "name"] == "Julia", c("age", "child")]
```
## Subsetting with matrix syntax

Excluding Cath, show `age` and `child` for people older than 28:

```{r}
people[people[, "name"] != "Cath" & 
       people[, "age"] > 28,
  c("name", "child")]
```

## Subsetting with matrix syntax

Filter out observations with missing age:

```{r}
people[!is.na(people[, "age"]), ]
```

## Subsetting with matrix syntax

Who's age is missing?

```{r}
people[is.na(people[, "age"]), "name"]
```

## Subsetting with matrix syntax

Show `name` and `child` for the people who are older than 21:

```{r}
people[people[, "age"] > 21, c("name", "child")]
```

Why do we have a row of `NA`s? 

## Subsetting with matrix syntax

```{r}
table(is.na(people$age))
```

Solution:

```{r}
people[!is.na(people[, "age"]) & people[, "age"] > 21,
       c("name", "child")]
```

## Subsetting with matrix syntax

Another solution:

```{r}

people_with_age <- people[!is.na(people[, "age"]), ]

people_with_age[people_with_age[, "age"] > 21,
  c("name", "child")]

```

## Subsetting with matrix syntax

If available, row names can be used for subsetting:

```{r}
df4

df4["Bob M.", ]
```

## Subsetting with matrix syntax

```{r}
df4

df4[c("Bob M.", "Sam L."), "age"]
```

## Subsetting with list syntax

```{r}
people
```

## Subsetting with list syntax

```{r}
people[2] # data frame
people["age"] # data frame

```

## Subsetting with list syntax

```{r}
people[c(1, 3)] # data frame
people[c("name", "child")] # data frame
```

## Subsetting with list syntax

```{r}
people["age"] # data frame
people[["age"]] # vector
people$age # vector
```

## Subsetting with list syntax

```{r}
people[["age"]][1]
people[["age"]][c(1, 4)]
people$age[1]
people$age[c(1, 4)]
```

## Subsetting with list syntax

```{r}
people[[2]] # same as people[["age"]]
people[[2]][1]
people[[2]][c(1, 4)]
```

## Subsetting with list syntax

Since data frames are lists of vectors, when you subset a data frame like a 1D object:

  - with `[` you get a data frame.
  - with `[[` or `$` you get a vector.

## Partial matching with `$`

When you attempt to extract a column from `df` with `df$name` and there is no column `name`, a data frame will instead select any variable that starts with `name`. If no variable starts with `name` (or if more than one do), `df$name` will return `NULL`.

```{r}
names(df)
df$a
```

By default, `[[` does not do partial matching. Why? Run `?"[["` and check what is the default value of the `exact` argument.

`df$name` is equivalent to `df[["name", exact = FALSE]]`

## Adding new columns

So far our data frame `people` has 5 rows and 3 columns.

```{r}
str(people)
```

## Adding columns with `[` 

Let’s add a column with people’s height:

```{r}
heights <- c(175, 170, 166, 182, 172) 
people["height"] <- heights

str(people)
```

## Adding columns with `[` 

We can use the position of the new column instead of the name:

```{r}
has_dog <- factor(c("yes", "no", "yes", "no", "no"))
people[5] <- has_dog

str(people)
```

## Adding columns with `[` 

What is the consequence of using the position of the new column instead of the name? 

  - Hint: look at the names of the new columns.
  
Lets rename `V5`:
  
```{r}
names(people)[5] <- "dog"

people
```

## Adding columns with `$`

Let’s add a column with people’s weight using `$`:

```{r}
weight <- c(86, 63, 68, 55, 56)
people$weight <- weight

str(people)
```

With `$` we can only add new columns by name.

## Adding columns with `cbind()`

Now let’s add a column with the people’s income:

```{r}
income <- c(1200, 750, 1660, 1280, 890) 
people <- cbind(people, income)
 
 str(people)

```

## Dropping columns with list syntax

Let's drop the `dog` column. By name:

```{r}
people$dog <- NULL 
# Or: people["dog"] <- NULL

people
```

## Dropping columns with list syntax

Now let's drop `income`. By position:

```{r}
people[6] <- NULL 
# or: people <-   people[-6]

people
```

## Reordering columns

To reorder columns, just select them in the desired order. For example, let's swap the `height` and `weight` columns:

```{r}
people <- people[, c("name", "age", "child",  "weight", "height")]

people
```

## Modifying columns

We can overwrite existing columns. Lets convert `height` from centimeters to meters:

```{r}
people$height <- people$height/100

people
```

## Modifying columns

In the previous example, the following commands would also have worked:

```{r, eval = FALSE}
people["height"] <- people["height"]/100
people[5] <- people[5]/100
```

## Calculations with existing columns

We can create new columns using the values of existing columns. Let’s use `weight` and `height` to calculate the body mass index:

```{r}
people$bmi <- round(
  people$weight/((people$height)^2)
  )
people
```

## Calculations with existing columns

Now let’s the the body mass index to calculate the weight status:

```{r}
people$weight_status <- ifelse(
  people$bmi < 18.5, "underweight",
    ifelse(people$bmi < 24.9, "normal", 
           ifelse(people$bmi < 29.9, "overweight", 
                  "obese")
           )
  )
```

## Calculations with existing columns

```{r}
people
```

## Dropping columns using matrix syntax

Let's drop the `bmi` and `weight_status` columns By name:

```{r}
people[, c("bmi", "weight_status")] <- NULL 
```

Now let's drop `weight`:  
  
```{r}
people[, 4] <- NULL 
# or people <- people[, -4] 
```
## Add a row 

```{r, error = TRUE}
tom <- data.frame("Tom", 37, FALSE, 1.83)
```

```{r, eval = FALSE}
rbind(people, tom)
## Error in match.names(clabs, names(xi)) : names do
## not match previous names
```

## Add a row 

The default names do not match the names of the `people` data frame:

```{r}
names(tom)
names(people)
```

## Add a row 

```{r}
tom <- data.frame(name = "Tom", age = 37, 
                  child = FALSE, height = 1.83)
rbind(people, tom)
```

## Add a row

You can also add rows using vectors, as long as they have appropriate names and the correct length:

```{r}
new_obs <- t(c(name = "Peter", age = 30, 
               child = FALSE, height = 1.86))

rbind(people, new_obs)
```

## Combining matrix and list syntax

Show the name and age of the people without children:

```{r}
people[people$child == FALSE, c("name", "age")]
```

## Combining matrix and list syntax

Show the name and age of the people who are taller than 1.70:

```{r}
people[people$height > 1.70, c("name", "age")]
```

## Combining matrix and list syntax

Show the name and age of the people without children who are taller than 1.70:

```{r}
people[people$child == FALSE & people$height > 1.70,
       c("name", "age")]
```

## Combining matrix and list syntax

Show the name, age and height of the people with children who are taller than 1.70:

```{r}
people[people$child == TRUE & people$height > 1.70,
       c("name", "age", "height")]
```

## Sorting

```{r}
people$age
sort(people$age)
ranks <- order(people$age)
ranks
people$age[ranks]

```

## Sorting

Sort the rows by ascending `age`:

```{r}
people[ranks, ]
```

## Sorting

Sort the rows by descending `age`:

```{r}
people[order(-people$age), ]
```

## Sorting

Sort the rows by descending `height`:

```{r}
people[order(people$height, decreasing = TRUE), ]
```

## Sorting

Sort the rows by `name`:

```{r}
people[order(people$name), ]
```

## The `head()` and `tails()` functions

- `head()` shows the first rows of a dataframe. 

- `tail()` shows the last rows. 

- Both `head()` and `tails()` print six rows by default. 

## The `head()` and `tails()` functions

```{r}
nrow(airquality)
head(airquality)
```

## The `head()` and `tails()` functions

```{r}
head(airquality, n = 2)
tail(airquality, n = 3) 
```

## The `head()` and `tails()` functions

`head()` and `tails()` also work with vectors and matrices:

```{r}
vec <- 1:100
mat <- matrix(1:100, ncol = 5, byrow = TRUE)
```

```{r}
head(vec)
tail(mat, n = 2)
```




