---
title: "Introduction to R Programming"
subtitle: "Getting Started"
author: Pedro Fonseca
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
output: 
  beamer_presentation:
    includes: 
      in_header: files/preamble.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R and Rstudio

- R is a programming language and a free and open source software environment 
for statistical computing and graphics

- RStudio is an integrated development environment (IDE) for R with free/open 
source and commercial versions

- You can use R without using RStudio, but you cannot use Rstudio without 
using R

## This is how R looks like

```{=tex}
\begin{figure}
\includegraphics[scale = .3]{figures/Screenshot_1}
\caption{R on Windows}
\end{figure}
```

## This is how R looks like

```{=tex}
\begin{figure}
\includegraphics[scale = .3]{figures/r-mac}
\caption{R on macOS}
\end{figure}
```

## This is how R looks like

```{=tex}
\begin{figure}
\includegraphics[scale = .3]{figures/r-linux}
\caption{R on Ubuntu}
\end{figure}
```

## Why use RStudio?

- R is command line driven

- RStudio provides more a user-friendly and interactive interface to R

## This is how Rstudio looks like

```{=tex}
\begin{figure}[H]
\includegraphics[scale = 0.17]{figures/Screenshot_mac}
\caption{RStudio}
\end{figure}
```
## The console

The pane on the left is the console. It can be used as a calculator:

```{=tex}
\begin{figure}[H]
\includegraphics[scale = 0.35]{figures/calc}
\caption{R's console as a calculator}
\end{figure}
```

## R as a calculator

```{r}
2 + 3 
3 * 5
14.5 / 6
3 ^ 2
```

## R as a calculator

We can chain as many operations as we want. But be careful with the parentheses!

```{r}
(3 ^ 2) + 14 / (6 + 5)
(3 ^ 2) + 14 / 6 + 5
```

## R as a calculator

Square root:

```{r}
sqrt(x = 25)
```

Natural logarithm:

```{r}
log(x = 5)
```

Common logarithm:

```{r}
log10(x = 5)
```

## R as a calculator

Exponential function:

```{r}
exp(x = 1)
exp(x = 3)
```

## R as a calculator

Nested operations:

```{r}
10 ^ log10(x = 5)
log(x = exp(x = 4))
sqrt(x = 25) ^ 2 + log(x = exp(x = 5))
```

## R as a calculator

Trigonometric functions:

```{r}
pi
cos(x = 2 * pi)
tan(x = 0.6)
sin(x = 0.6) / cos(x = 0.6)
```

## Functions

- R includes a large collection of built-in functions

- We already used `log()`, `log10()`, `sqrt()`, `exp()`, `sin()`, `cos()` and
`tan()`

## Functions

Most functions have more than one argument:

- Some arguments are mandatory

- Some arguments are optional and have default values

## Functions

The `log()` function has two arguments:

- `x` is mandatory

- `base` is optional. The default value of `base` is `exp(1)`

```{r}
log(x = 243)
log(x = 243, base = exp(1))
log(x = 243, base = 2)
```

## Functions

The `log10()` function only has one argument, `x`, and it is mandatory:

```{r}
log10(x = 243)
log(x = 243, base = 10)
```

## Functions

The `round()` function rounds numbers to a specified number of decimal places. 
It has two arguments:

- `x` is mandatory

- `digits` is optional. The default value of `digits` is `0`

```{r}
round(x = 5.23452)
round(x = 5.23452, digits = 2)
round(x = 5.23452, digits = 3)
```

## Functions

`sqrt()`, `log10()`, `exp()`, `sin()`, `cos()` and `tan()` have only one
argument, `x`, and it is mandatory:

```{r}
sqrt(x = 5)
log10(x = 5)
exp(x = 5)
```

## Functions

Argument names are not mandatory:

```{r}
log(x = 5, base = 10)
log(x = 5, 10)
log(5, base = 10)
log(5, 10)
```

## Functions

Dropping the names of the arguments is safe in functions with only one argument:

```{r}
sqrt(x = 25)
sqrt(25)
```

## Functions

R does positional matching for unnamed arguments. Therefore, in functions with 
more than one argument we must pay attention to the ordering of the arguments:

```{r}
log(243, 2)
log(2, 243)
```

## Functions

If we provide the names of the arguments, the ordering is irrelevant:

```{r}
log(x = 243, base = 2)
log(base = 2, x = 243)
```

## Functions

It is usually safe to drop the name of the first argument. Providing the names 
of the remaining arguments is usually a good idea: it avoids mistakes and 
improves readability.

```{r}
log(4, base = 3)
round(pi, digits = 2)
round(sqrt(2), digits = 4)
```

## Functions

Help pages can be very useful:

```{r, eval=FALSE}
?log
```

In the help page of a function you can find:

- An ordered list of the function's arguments

- Details about the arguments and their admissible values

- Details about how the function works

- The interpretation of the output of the function

- Examples

- Related functions

## R scripts

- So far we have only been using Rstudio's console

- Code sent directly to the console is executed but you won't be able to modify 
it or reuse it later

- Using scripts is a better option

- A script is just a text file that we can use to write code

## Your first R script

```{=tex}
\begin{figure}
\includegraphics[scale=0.35]{figures/new-script}
\caption{Creating a new script}
\end{figure}
```

## Editor

- R opens scripts in the editor pane (top left)

- This is where you should write your code

- In the editor you can modify, rerun and save your code at any time

- Scripts are the holly grail of reproducible data analysis

## Rstudio Panes

```{=tex}
\begin{figure}
\includegraphics[scale=0.18]{figures/panes}
\caption{R studio's console and editor}
\end{figure}
```
## Some useful shortcuts

- New script: Cmd/Ctrl + Shift + N

- Save the script: Cmd/Ctrl + S

- Send code from the script to the console:

  - Cmd/Ctrl + Enter (current line or current selection)

  - Cmd/Ctrl + Shift + S (entire script)

## The assignment operator

To store values in R's memory you need to assign them to objects. You can use 
the equal sign (`=`) or the assignment operator (`<-`):

```{r}
x <- 5
x
```

## The assignment operator

- The assignment operator is a better option

- The equal sign should be reserved to provide arguments to functions

- Rstudio's shortcut to the assignment operator is "Alt/Option" + "$-$"

## The assignment operator

\includegraphics[scale=.22]{figures/meme}

## The assignment operator

Values stored in objects can be used in calculations:

```{r}
y <- log(x) + exp(2) 
x + 2 * y
```

## The assignment operator

Stored objects are visible in the upper-right pane, under the "Environment" tab:

```{=tex}
\begin{figure}
\includegraphics[scale=.45]{figures/objects}
\caption{Our session's global envoronment}
\end{figure}
```

## Workflow

Our workflow so far:

- Write code in the editor

- Send code to the console

- The code is exectued and the results are printed in the console

- The objects we created are listed in the environment tab

## Workflow

```{=tex}
\begin{figure}
\includegraphics[scale=0.18]{figures/wf}
\caption{Workflow}
\end{figure}
```
## Commenting

- We can make comments in our code using `#`

- Lines starting with `#` are printed in the console but are not executed

## Commenting

```{r}
#===================================================
# Intro to R programming - Lecture 0
#===================================================

# Lets store the value "5" in an object called x
x <- 5 

# Now let's print x
x
```

## Saving scripts

Since we already edited our script, let's save it:

```{=tex}
\begin{figure}
\includegraphics[scale=0.32]{figures/save}
\caption{Saving a script}
\end{figure}
```

## Saving scripts

```{=tex}
\begin{figure}
\includegraphics[scale=0.18]{figures/saved}
\caption{The script is now saved}
\end{figure}
```

## Naming Objects

- Object names must start with a letter and can only contain letters, numbers, 
underscores and dots

- Ideally, one should follow a convention

- Object names should to be short, descriptive and consistent

## Case matters

```{r}
r_rocks <- 2 
r_rocks
```

```{r, eval = FALSE}
r_Rocks
```

`## Error: object 'r_Rocks' not found`

## How to delete objects

To delete stored objects use the `rm` function:

```{r}
r_rocks

rm(r_rocks)
```

```{r, eval = FALSE}
r_rocks
```

`## Error: object 'r_rocks' not found`

## How to delete objects

You can input as many objects as you want to `rm()`:

```{r}
rm(x, y)
```

To remove all stored objects all once, use the following command:

```{r, eval=FALSE}
rm(list = ls())
```

<!-- ## How to print the assigned value -->

<!-- Assignment is a silent operation: when you make an assignment, you don't 
get to see the assigned value. You're then tempted to double-check the result: 
-->

<!-- ```{r} -->

<!-- y <- log(2) + 1 -->

<!-- y -->

<!-- ``` -->

<!-- This common action can be shortened by surrounding the assignment with 
parentheses, which causes assignments to print: -->

<!-- ```{r} -->

<!-- (y <- log(2) + 1) -->

<!-- ``` -->

## Overwritting stored values

```{r}
x <- -5
x

x <- x + 1 
x

x <- round(log(3)/2, digits = 2)
x
```

## Working directory

An R session always has an associated working directory. R will use the working 
directory by default to:

- Search for files

- Save files

- Save outputs (tables, plots, etc)

To check your working directory:

```{r, eval = FALSE}
getwd()
```

## Setting the working directory

You can change your working directory using RStudio's menus:

```{=tex}
\begin{figure}
\includegraphics[scale = .40]{figures/wd}
\caption{Setting the working directory}
\end{figure}
```

## Setting the working directory

```{=tex}
\begin{figure}
\includegraphics[scale = .35]{figures/wd2}
\caption{Setting the working directory}
\end{figure}
```

## Setting the working directory

You can also change the working directory in R's console:

```{r, eval=FALSE}
setwd("/Users/pedro/Documents/intro-to-r")
```

## Setting the working directory

- The problem with absolute paths like the one in the last slide is that they 
only exist in my computer

- This makes it more difficult to share and reproduce scripts

- Solution: Rstudio projects

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.31]{figures/new-project-1}
\caption{Creating a new project}
\end{figure}
```

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.47]{figures/new-project-2}
\caption{Creating a new project}
\end{figure}
```

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.47]{figures/new-project-3}
\caption{Creating a new project}
\end{figure}
```

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.36]{figures/new-project-4}
\caption{Creating a new project}
\end{figure}
```

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.5]{figures/new-project-5}
\caption{Creating a new project}
\end{figure}
```

## Your first Rstudio project

```{=tex}
\begin{figure}
\includegraphics[scale=0.19]{figures/environment}
\caption{Creating a new project}
\end{figure}
```

## Advantages of Rstudio projects

- Rstudio projects are self-contained.

- They put together all the files that are relevant for a particular project 
(article, book, research project) in the same folder

- The project's working directory always points to that folder by default

- Rstudio projects can be moved around on your computer or onto other computers 
and will still "just work". No directory changes are needed.

- If you need to create additional folders or start moving around parts of you 
project around dont use the `setwd` function. It is safer to reference the full 
path.

## Packages

- The more specialized functions are distributed on packages

- Packages are developed by the R core team and also by the community of R users

- You can develop your own packages and make them available to the community 
through [CRAN](https://cran.r-project.org) (The Comprehensive R Archive Network)

## Packages

Later in this course, we will use the `sqldf` package. Let's install it:

```{r, eval=FALSE}
install.packages("sqldf")
```

If you want to use an installed package, you must load it first:

```{r, eval=FALSE}
library("sqldf")
```

Update an installed package:

```{r, eval=FALSE}
update.packages("sqldf") 
```

## Packages

- It is recommended that you start your scripts by loading the packages that 
will be used

- That way, if you share your code with others (even if that's future you), they 
can easily see what packages they need to install

## Packages

- Note, however, that you should never include `install.packages` or `setwd` in 
a script that you share

- Use `library` instead

- It is very antisocial to change settings or install software on someone else's 
computer!
