---
title: "Introduction to R Programming"
subtitle: "Data Frames"
author: Pedro Fonseca
date: "`r format(Sys.time(), '%d %B %Y')`"
urlcolor: blue
output: beamer_presentation
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The `c()` Operator

- Vectors are essential building blocks for handling multiple items in R.  

- To create vectors use the *combine* operator:

```{r}

x <-  -1

y <-  10

z <-  c(1, 6, 3)

(myvec <- c(1, 3, 1, 42))

(myvec2 <- c(myvec, x, y ,z))

(myvec3 <-  c(myvec, 1, 2))
```

## Subsetting

Get the first element:

```{r}
myvec[1]
```

Get the second element:

```{r}
myvec[2]
```


## Subsetting

Get the first three elements:

```{r}
myvec[1:3]
```

Omit the first element:

```{r}
myvec[-1]
```

Omit more than one element:

```{r}
myvec[-c(1,2)]
```

## Overwriting

Substitute an element:

```{r}
myvec[3] <- 6
myvec
```

Substitute more than one element:

```{r}
myvec[c(2,3,4)] <- c(2,3,4)

```

## Functions to Generate Vectors

```{r}
1:10
5:1
seq(1, 10)
seq(from = 18, to = 27, by = 3)
```


## Functions to Generate Vectors

```{r}
rep(x = 1, times = 4)
rep(c(3, 6), times = 3)
rep(c(3, 62, 8.3), each = 2)
rep(c(3, 6), times = 3, each = 2)
```

## Sorting the Elements of a Vector


Sorting a vector in increasing or decreasing order: 

```{r}
myvec2 <- c(1, 3, 1, 42, -5, 10, -50)
myvec2
sort(myvec2)
sort(myvec2, decreasing = TRUE)
```


## Sorting the Elements of a Vector

```{r}
sort(c(2.5, -1, -10, 3.44))
sort(c(2.5,-1,-10,3.44), decreasing = TRUE)
sort(c(2.5,-1,-10,3.44), TRUE)
```

## Some Statistical Functions

`rnorm(n)` generates `n` pseudo-random numbers from a normal distribution (default: $\mu = 0$, $\sigma = 1$)

```{r}
rnorm(3)
rnorm(4, mean =  5, sd = 2)
```

Other functions related do the normal distribution:

  - `dnorm` (density), `pnorm` (distribution function), `qnorm` (quantile function).
  
  - Equivalent functions are available for the most commonly used probability distributions: F, t-student, Uniform, Poisson...

## The `set.seed` Function

- Functions like `rnorm`, `rpois` and `runif` generate pseudo-random numbers. 

- This means that you and I will get different results when using these functions. 

- Solution: use the `set.seed` function.

## The `set.seed` Function

Try this command many times: 

```{r}
rnorm(2) 
```

Each time you will get a different output. Now try this:

```{r}
set.seed(123)
rnorm(2)
```

You will get the  same output every time. The argument of \textit{set.seed} is irrelevant as long as we all use the same value.

## Vectorized Operations

One of the main advantages of R is vectorized calculation. This means that:

- Most R functions accept vectors as inputs;

- Vector arithmetic is performed element-wise by default.

## Vectorized Operations

- Vectorization calculation is a huge advantage efficiency and parsimony.

-  Vectorization also makes code easier to write and read.

## Vectorized Operations

```{r}
x <- c(1, 2, 3)
y <- c(0.5, 0.5, 0.5)
1/x
3+y
x+y

```

## Vectorized Operations

```{r}
x <- c(1, 2, 3)
y <- c(0.5, 0.5, 0.5)

x^y
sqrt(x)
1/1:3
seq(from = 2, to = 6, by = 2)/2

```

## Vectorized Operations

```{r}
x1 <- c(1, 5, 7)
x2 <- rep(1, times = 3)

log(x1)
log(x1) - x2

x <- x1 + x2
x
```

## Rounding 

`round()` rounds the values in its first argument to the specified number of decimal places (default 0):

```{r}
set.seed(123)

z <- rnorm(3)
z

round(z, digits = 3)
round(z)
```

## Rounding 

```{r}
y <- c(3.271109, 3.374961, 2.313307, 4.837787)
round(y, 2)
```
 
## Statistical Functions

```{r}
z
abs(z) # Absolut value
max(z)
min(z)
```

## Statistical Functions

```{r}
z

mean(z)
median(z)
sd(z)
```

## Statistical Functions

```{r}
z
var(z)
sum(z)
quantile(z, 0.5)
```

## The `which()` Function

The `which()` function is useful to find which elements of a vector that verify a given condition:

```{r}
set.seed(123)
vec <- rnorm(n = 10, mean = 2, sd = 1)
round(vec, 2)
(indexes <- which(vec >  2))
round(vec[indexes], 3)
```

## The `which()` Function

```{r}
set.seed(123)

vec2 <- rpois(n = 10, lambda = 2)
which(vec2 == 2)
(vec2 <- rpois(n = 10, lambda = 2)) 
which(vec2 == 2)
```

## The `which()` Function

```{r}
set.seed(123)

vec2 <- rpois(n = 10, lambda = 2)
vec2
max(vec2)
which(vec2 == max(vec2))

```

## The `which()` Function

The `which` function gives the positions of the elements of the vectors that verify the condition, not their values!

```{r}
set.seed(123)

vec2 <- rpois(n = 10, lambda = 2)
vec2
```

What are the actual values of \textit{vec2} (not their positions) that verify the condition?

```{r}
vec2[which(vec > 1)]
```

## The `length()` Function

```{r}
round(vec[which(vec > 2)], 3)
```

Use `length()` to obtain the number of elements in a vector:

```{r}
length(vec)
```

How many elements of `vec` are greater than 2?

```{r}
length(which(vec > 2))
```

## Trigonometric Functions

R trigonometric take radians as argument, not degrees:

- $\operatorname{sin}(\frac{\pi}{2})$:

```{r}
sin(pi/2)
```

- $\operatorname{cos}(\pi)$:

```{r}
cos(pi)
```

- $\operatorname{tan}(\frac{\pi}{3})$:

```{r}
tan(pi/3)
```

- $\operatorname{cotangent}(\frac{\pi}{3})$:

```{r}
1/tan(pi/3)
```

## Trigonometric Functions

Which value has a cosine = -1?

```{r}
acos(-1)
cos(pi)
```

Which value has a tangent = 0.5? 

```{r}
atan(0.5)
tan(0.4636476)
```

## Trigonometric Functions

Trigonometric functions are also vectorized:

```{r}
(x <- seq(from = 0.25, to = 1, by = 0.25))
cos(x)
1/tan(x) # cotangent of x
cos(x)/sin(x) # cotangent of x
```

## Trigonometric Functions

R has many more trigonometric functions. Try:

```{r}
?Trig
```

## Recycling

What happens when we conduct calculations with two vectors of different length?

```{r}
myvec <- c(1, 2, 3, 4) 
myvec2 <- rep(0.5, times = 8)

myvec + myvec2
```

## Recycling

```{r, warning = TRUE}
myvec3 <- rep(0.5, times = 7)

myvec + myvec3
```

## Recycling

- When conducting operations that require input vectors to be of the same length, R automatically recycles, or repeats, the shorter one, until it is long enough to match the longer one.

- It will only throw an error message if the length of the shorter vector is not a multiple of the vector of the larger vector.

## Named Vectors

We can also name the elements of a vector:

```{r}
x <- c(x1 = 1, x2 =  4, x3 = 7)
x
```

Get the names of a vector:

```{r}
names(x)
```

## Named Vectors

The `names()` function can also be used to provide names to a vector:

```{r}
y <- 1:3
names(y) <- c("y1", "y2", "y3")
```

## Subsetting Named Vectors

Vectors can also be subseted by name:

```{r}
y
y["y1"]
y[c("y1", "y3")]
```


## The `paste()` and `paste0()` functions

`paste()` and `paste0()` can be useful to generate vector names:

```{r}
paste("y", 1:length(y), sep = "")
paste("name", 1:length(y), sep = "_")
paste("year", 1990:1993, sep = "-")
paste0("X", 1:5)
```

## Built in constants

R provides some built in constants and vectors. We already used one:

```{r}
pi
```

## Built in constants

In this lecture we will also use `letters`:

```{r}
head(letters)
tail(letters)
```

## Built in constants

There is also an upper-case version:

```{r}
head(LETTERS)
tail(LETTERS)
```

For a complete list of built in constants see `?Constants`

